Characterizing Signal Loss, Error, and Bias in the 21cm Reionization \\Power Spectrum: A Revised Study of PAPER-64

### 2018-06-07 PAPER.tex
changed dC/dP's to Q's and changed some minor syntax stuff in Section 2.1.1

The goal of power spectrum analysis is to produce an unbiased estimator of the EoR power spectrum in the presence of both noise and foreground emission. Prior to power spectrum estimation, the data will often have been prepared to have minimal foregrounds by some method of subtraction, so this foreground emission may appear either directly (because it was not subtracted) or as a residual of some subtraction process not in the power spectrum domain. If an accurate estimate of the total covariance of the data is known, including both the desired signal and any contaminants, then the ``optimal quadratic estimator'' formalism provides a method of producing a minimum variance, unbiased estimator of the desired signal, as shown in 
% JEA - I cut this b/c I wanted to step back and explain a bit more clearly what we were actually hoping for 
%Driven by the need to mitigate foreground bias, PAPER's previous analyses use a weighting method that aims to down-weight foregrounds. This weighting is applied to data, which is then propagated into a final estimator using the power spectrum estimation technique of quadratic estimators (QE) as done in 
\citet{liu_tegmark2011}, \citet{dillon_et_al2013a}, \citet{liu_et_al2014a}, \citet{liu_et_al2014b}, \citet{trott_et_al2012}, \citet{dillon_et_al2014}, \citet{dillon_et_al2015}, \citet{switzer_et_al2015}, and \citet{trott_et_al2016}. 
%Before showing how signal loss can arise when using different weighting matrices, we first summarize QE as performed in the PAPER analysis.
We provide a brief summary of the optimal quadratic estimator below.

@@

We are interested in estimating the three-dimensional power spectrum of the EoR.  
Visibilities are measurements of the Fourier transform of the sky along two spatial dimensions (using the flat-sky approximation), and since we are interested in three-dimensional Fourier modes we only need to take one Fourier transform of our visibilities along the line-of-sight dimension.  We consider band powers $p_\alpha$ of the power spectrum of $\textbf{x}$ over some range in cosmological $\mathbf{k}$, where $\alpha$ indexes a waveband in $k_{\parallel}$ (a cosmological wavenumber $k_{\parallel}$ is the Fourier dual to frequency under the delay approximation (\citealt{parsons_et_al2012b}), which is a good approximation for the short baselines that PAPER analyzes).  The fundamental dependence of the covariance on the power spectrum band powers $p_\alpha$ is encoded as 
% JEA - Should distinguish slightly more clearly between Q and dC/dp_alpha
\begin{equation}
\textbf{S} = p_{\alpha} \frac{\partial\textbf{C}}{\partial p_{\alpha}}
%\frac{\partial\textbf{C}}{\partial p_{\alpha}} \equiv \textbf{Q}^{\alpha}
\end{equation}
where we use the Einstein summation convention.
% JEA - ah, the dimensions again
% which has dimensions ($N_{f}$,$N_{f}$).

The optimal quadratic estimator prescription is then to compute
\begin{equation}
\label{eq:OQE}
\widehat{p}_{\alpha}  =  ({\F^{-1}})_\alpha^{\beta} (\widehat{q}_{\beta} - \widehat{b}_{\beta} )
\end{equation}
where $\F$ is the Fisher matrix 
\begin{equation}
F^{\alpha \beta} \equiv \frac{1}{2} \textrm{tr} \left( \C^{-1} \frac{\partial\textbf{C}}{\partial p_{\alpha}} \C^{-1} \frac{\partial\textbf{C}}{\partial p_{\beta}} \right)
\end{equation}
and
\begin{equation}
\label{eq:OQEQuadratic}
\hat{q}_\alpha =  \half \textbf{x}^\dagger \invC \PDeriv{\C}{p^\alpha}  \invC \textbf{x} 
\end{equation}
and
\begin{equation}
\label{eq:OQELinear}
\widehat{b}_{\alpha} = \half \Tr\left( \mathbf{U} \invC \PDeriv{\C}{p^\alpha} \invC \right).
\end{equation}
This estimator is the minimum variance (smallest error bar) estimate of the power spectrum  subject to the constraint that it is also unbiased; that is, the ensemble average of the estimator is equal to its true value
\begin{equation}
\label{eq:super_unbiased}
\langle \widehat{p}_{\alpha} \rangle = p_\alpha.
\end{equation}
\citep{tegmark_et_al1997a,bond_et_al1998}

It is clear that in order to obtain an unbiased estimate in the presence of contaminants, the estimator defined by Equation \eqref{eq:OQE} must be capable of ``suppressing'' or ``removing'' their effects. By construction, the subtraction of the residual foreground and noise bias accomplishes this, removing any additive bias. However, the $\mathbf{C}^{-1}$ piece of Equation \eqref{eq:OQEQuadratic} also has the effect of suppressing residual foregrounds and noise, in both the additive bias and any contributions the residuals may have to the variance. The way in which the optimal estimator accomplishes this is illustrated with a toy model in Appendix \ref{sec:icw_appendix}. In this appendix, we show that the effect of the weighting in Equation \ref{eq:OQEQuadratic} is to project out the modes of $\textbf{U}$ with a different covariance structure than $\mathbf{S}$ in the power spectrum estimate, and the effect of Equation \ref{eq:OQELinear} is to subtract out the remaining bias. Similar effects for a more realistic model of the EoR and foregrounds are shown in \citet{liu_tegmark2011}.

The toy model in Appendix \ref{sec:icw_appendix} also illustrates that if the covariance structure of the contaminants is sufficiently different from the desired power spectrum, then the linear bias term may be expected to be quite small, and it is only necessary to know $\C$ and $\partial \C/\partial p_\alpha$, but not $\mathbf{U}$. Since the foregrounds are expected to be strongly correlated between frequencies whereas the EoR is not, we expect different covariance structures and therefore a small linear bias. Moreover, because the linear bias is always positive and there is no multiplicative bias, the quadratic-only term will always produce an estimate which is {\it high} relative to the true value, and which can conservatively be interpreted as an upper limit. These considerations, and the difficulty of obtaining an estimate for $\mathbf{U}$, motivate the neglect of the linear bias in the rest of this analysis.

%We do this when forming the un-normalized power spectrum estimate $\widehat{q}_{\alpha}$:
Motivated by the desire to retain the desirable behavior of suppressing contributions of $\mathbf{U}$ to estimates of the EoR power spectrum, we note that is possible to define a modified version of the quadratic estimator 
%which may still produce an unbiased estimate, 
where Equation \eqref{eq:OQEQuadratic} is replaced by
\begin{equation}
\label{eq:qhat}
\widehat{q}_{\alpha} = \frac{1}{2}\textbf{x}^{\dagger}\textbf{R}\textbf{Q}^{\alpha}\textbf{R}\textbf{x}.
\end{equation}
where $\textbf{R}$ is a weighting matrix chosen by the data analyst.  For example, inverse covariance weighting (the optimal form of QE) would set $\textbf{R} \equiv \textbf{C}^{-1}$ and a uniform-weighted case would use $\textbf{R} \equiv \textbf{I}$, the identity matrix.
The matrix $\textbf{Q}^{\alpha}$ encodes the dependence of the covariance on the power spectrum parameters $\partial \C/\partial p_\alpha$ but in practice also does other things, including implementing a transform of the frequency-domain visibilities to $\mathbf{k}$-space, taking into account cosmological scalings, and converting the visibilities from Jy to Kelvin.
%with dimensions ($N_{f}$,$N_{f}$) --- as

%%% COME BACK HERE
% Need to still explain the motivation for this
With an appropriate normalization matrix $\textbf{M}$, the quantity
%We normalize our power spectrum estimates using the matrix $\textbf{M}$:
\begin{equation}
\label{eq:phat}
\widehat{\textbf{P}} = \textbf{M}\widehat{\textbf{q}}
\end{equation}
is a sensible estimate of the true power spectrum $\textbf{P}$. To ensure that $\textbf{M}$ correctly normalizes our power spectrum, one may take the expectation value of Equation \eqref{eq:phat} to obtain
\begin{eqnarray}
\langle \widehat{P}_\alpha \rangle &=& \frac{1}{2} \sum_{\beta \gamma} M_{\alpha \gamma}  \textrm{tr} \left( \textbf{R}\textbf{Q}^{\gamma}\textbf{R} \textbf{Q}^{\beta}\right) p_\beta +\frac{1}{2} \sum_{ \gamma}  \Tr\left( \mathbf{U} \textbf{R} \textbf{Q}^{\gamma} \mathbf{R} \right) \nonumber \\
&\equiv & \sum_{\beta} W_{\alpha \beta} p_\beta +\frac{1}{2} \sum_{ \gamma}  \Tr\left( \mathbf{U} \textbf{R} \textbf{Q}^{\gamma} \mathbf{R} \right), \label{eq:Wpplusbias}
\end{eqnarray}
where $W_{\alpha \beta}$ are elements of a window function matrix. Consider the first term of this expression. (Again, we are assuming that the linear bias term is significantly suppressed; and if this is not the case, we are simply assuming that we are setting a conservative upper limit). If $\mathbf{W}$ ends up being the identity matrix for our choices of $\mathbf{R}$ and $\mathbf{M}$, then we recover Equation \eqref{eq:super_unbiased} for the first term, and we have an estimator that has no multiplicative matrix bias. However, Equation \eqref{eq:super_unbiased} is a rather restrictive condition, and it is possible to violate it and still have a sensible (and correctly normalized) power spectrum estimates. In particular, as long as the rows of $\mathbf{W}$ sum to unity, our power spectrum will be correctly normalized. Beyond this, the data analyst has a choice for $
\textbf{M}$. For simplicity in this section we choose $\textbf{M}$ to be diagonal, although we explore other cases for the analysis 
of PAPER-64 data as explained in Section \ref{sec:Bias}. In a preview of what is to come, we also stress that the derivation that leads to Equation \eqref{eq:Wpplusbias} assumes that $\mathbf{R}$ and $\mathbf{x}$ are not correlated. If this assumption is violated, a simple application of the (now incorrect) formulae in this section can result in an improperly normalized power spectrum estimator that does not conserve power, i.e., one that has signal loss.
% JEA - is one really free to choose M and R independently?

@@

As a brief summary, in the following sections, we find:

>>>>

The goal of power spectrum analysis is to produce an unbiased estimator of the EoR power spectrum in the presence of both noise and foreground emission. Prior to power spectrum estimation, the data will often have been prepared to have minimal foregrounds by some method of subtraction, so this foreground emission may appear either directly (because it was not subtracted) or as a residual of some subtraction process not in the power spectrum domain. If an accurate estimate of the total covariance of the data is known, including both the desired signal and any contaminants, then the ``optimal quadratic estimator'' formalism provides a method of producing a minimum variance, unbiased estimator of the desired signal, as shown in 
% JEA - I cut this b/c I wanted to step back and explain a bit more clearly what we were actually hoping for 
%Driven by the need to mitigate foreground bias, PAPER's previous analyses use a weighting method that aims to down-weight foregrounds. This weighting is applied to data, which is then propagated into a final estimator using the power spectrum estimation technique of quadratic estimators (QE) as done in 
\citet{liu_tegmark2011}, \citet{dillon_et_al2013a}, \citet{liu_et_al2014a}, \citet{liu_et_al2014b}, \citet{trott_et_al2012}, \citet{dillon_et_al2014}, \citet{dillon_et_al2015}, \citet{switzer_et_al2015}, and \citet{trott_et_al2016}. 
%Before showing how signal loss can arise when using different weighting matrices, we first summarize QE as performed in the PAPER analysis.
We provide a brief summary of the optimal quadratic estimator and explain its advantages below.

@@

We are interested in estimating the three-dimensional power spectrum of the EoR.  
Visibilities are measurements of the Fourier transform of the sky along two spatial dimensions (using the flat-sky approximation), and since we are interested in three-dimensional Fourier modes we only need to take one Fourier transform of our visibilities along the line-of-sight dimension.  We consider band powers $p_\alpha$ of the power spectrum of $\textbf{x}$ over some range in cosmological $\mathbf{k}$, where $\alpha$ indexes a waveband in $k_{\parallel}$ (a cosmological wavenumber $k_{\parallel}$ is the Fourier dual to frequency under the delay approximation (\citealt{parsons_et_al2012b}), which is a good approximation for the short baselines that PAPER analyzes).  The fundamental dependence of the covariance on the power spectrum band powers $p_\alpha$ is encoded as 
% JEA - Should distinguish slightly more clearly between Q and dC/dp_alpha
\begin{equation}
\textbf{S} = \sum_\alpha p_\alpha \frac{\partial\textbf{C}}{\partial p_\alpha} \equiv \sum_\alpha p_\alpha \textbf{Q}_\alpha
%\textbf{S} = p_{\alpha} \frac{\partial\textbf{C}}{\partial p_{\alpha}}
%\frac{\partial\textbf{C}}{\partial p_{\alpha}} \equiv \textbf{Q}^{\alpha}
\end{equation}
where we define $\frac{\partial\textbf{C}}{\partial p_\alpha} \equiv \textbf{Q}_{\alpha}$.
% JEA - ah, the dimensions again
% which has dimensions ($N_{f}$,$N_{f}$).

The optimal quadratic estimator prescription is then to compute
\begin{equation}
\label{eq:OQE}
\widehat{p}_{\alpha}  =  ({\F^{-1}})_\alpha^{\beta} (\widehat{q}_{\beta} - \widehat{b}_{\beta} )
\end{equation}
where $\F$ is the Fisher matrix 
\begin{equation}
F^{\alpha \beta} \equiv \frac{1}{2} \textrm{tr} \left( \C^{-1} \textbf{Q}^{\alpha} \C^{-1} \textbf{Q}^{\beta} \right)
\end{equation}
and
\begin{equation}
\label{eq:OQEQuadratic}
\widehat{q}_\alpha =  \half \textbf{x}^\dagger \invC \textbf{Q}^{\alpha}  \invC \textbf{x} 
\end{equation}
and
\begin{equation}
\label{eq:OQELinear}
\widehat{b}_{\alpha} = \half \Tr\left( \mathbf{U} \invC \textbf{Q}^{\alpha} \invC \right).
\end{equation}
This estimator is the minimum variance (smallest error bar) estimate of the power spectrum  subject to the constraint that it is also unbiased; that is, the ensemble average of the estimator is equal to its true value
\begin{equation}
\label{eq:super_unbiased}
\langle \widehat{p}_{\alpha} \rangle = p_\alpha.
\end{equation}
\citep{tegmark_et_al1997a,bond_et_al1998}

It is clear that in order to obtain an unbiased estimate in the presence of contaminants, the estimator defined by Equation \eqref{eq:OQE} must be capable of ``suppressing'' or ``removing'' their effects. By construction, the subtraction of the residual foreground and noise bias accomplishes this, removing any additive bias. However, the $\mathbf{C}^{-1}$ piece of Equation \eqref{eq:OQEQuadratic} also has the effect of suppressing residual foregrounds and noise, in both the additive bias and any contributions the residuals may have to the variance. The way in which the optimal estimator accomplishes this is illustrated with a toy model in Appendix \ref{sec:icw_appendix}. In this appendix, we show that the effect of the weighting in Equation \ref{eq:OQEQuadratic} is to project out the modes of $\textbf{U}$ with a different covariance structure than $\mathbf{S}$ in the power spectrum estimate, and the effect of Equation \ref{eq:OQELinear} is to subtract out the remaining bias. Similar effects for a realistic model of the EoR and foregrounds are shown in \citet{liu_tegmark2011}.

The toy model in Appendix \ref{sec:icw_appendix} also illustrates that if the covariance structure of the contaminants is sufficiently different from the desired power spectrum, then the linear bias term may be expected to be quite small, and it is only necessary to know $\C$ and $\textbf{Q}^{\alpha}$, but not $\mathbf{U}$. Since the foregrounds are expected to be strongly correlated between frequencies whereas the EoR is not, we expect different covariance structures and therefore a small linear bias. Moreover, because the linear bias is always positive and there is no multiplicative bias, the quadratic-only term will always produce an estimate which is {\it high} relative to the true value, and which can conservatively be interpreted as an upper limit. These considerations, and the difficulty of obtaining an estimate for $\mathbf{U}$, motivate the neglect of the linear bias in the rest of this analysis.

%We do this when forming the un-normalized power spectrum estimate $\widehat{q}_{\alpha}$:
Motivated by the desire to retain the advantageous behavior of suppressing contributions of $\mathbf{U}$ to estimates of the EoR power spectrum, we note that is possible to define a modified version of the quadratic estimator 
%which may still produce an unbiased estimate, 
where Equation \eqref{eq:OQEQuadratic} is replaced by
\begin{equation}
\label{eq:qhat}
\widehat{q}_{\alpha} = \frac{1}{2}\textbf{x}^{\dagger}\textbf{R}\textbf{Q}^{\alpha}\textbf{R}\textbf{x}
\end{equation}
where $\textbf{R}$ is a weighting matrix chosen by the data analyst.  For example, inverse covariance weighting (the optimal form of QE) would set $\textbf{R} \equiv \textbf{C}^{-1}$ and a uniform-weighted case would use $\textbf{R} \equiv \textbf{I}$, the identity matrix.
The matrix $\textbf{Q}^{\alpha}$ encodes the dependence of the covariance on the power spectrum parameters $\partial \C/\partial p_\alpha$ but in practice also does other things, including implementing a transform of the frequency-domain visibilities to $\mathbf{k}$-space, taking into account cosmological scalings, and converting the visibilities from Jy to Kelvin.
%with dimensions ($N_{f}$,$N_{f}$) --- as

%%% COME BACK HERE
% Need to still explain the motivation for this
With an appropriate normalization matrix $\textbf{M}$, the quantity
%We normalize our power spectrum estimates using the matrix $\textbf{M}$:
\begin{equation}
\label{eq:phat}
\widehat{\textbf{P}} = \textbf{M}\widehat{\textbf{q}}
\end{equation}
is a sensible estimate of the true power spectrum $\textbf{P}$. To ensure that $\textbf{M}$ correctly normalizes our power spectrum, one may take the expectation value of Equation \eqref{eq:phat} to obtain
\begin{eqnarray}
\langle \widehat{P}_\alpha \rangle &=& \frac{1}{2} \sum_{\beta \gamma} M_{\alpha \gamma}  \textrm{tr} \left( \textbf{R}\textbf{Q}^{\gamma}\textbf{R} \textbf{Q}^{\beta}\right) p_\beta +\frac{1}{2} \sum_{ \gamma}  \Tr\left( \mathbf{U} \textbf{R} \textbf{Q}^{\gamma} \mathbf{R} \right) \nonumber \\
&\equiv & \sum_{\beta} W_{\alpha \beta} p_\beta +\frac{1}{2} \sum_{ \gamma}  \Tr\left( \mathbf{U} \textbf{R} \textbf{Q}^{\gamma} \mathbf{R} \right), \label{eq:Wpplusbias}
\end{eqnarray}
where $W_{\alpha \beta}$ are elements of a window function matrix. Let's consider the first term of this expression (again, we are assuming that the linear bias term is significantly suppressed; and if this is not the case, we are simply assuming that we are setting a conservative upper limit). If $\mathbf{W}$ ends up being the identity matrix for our choices of $\mathbf{R}$ and $\mathbf{M}$, then we recover Equation \eqref{eq:super_unbiased} for the first term, and we have an estimator that has no multiplicative matrix bias. However, Equation \eqref{eq:super_unbiased} is a rather restrictive condition, and it is possible to violate it and still have a sensible (and correctly normalized) power spectrum estimate. In particular, as long as the rows of $\mathbf{W}$ sum to unity, our power spectrum will be correctly normalized. Beyond this, the data analyst has a choice for $
\textbf{M}$. For simplicity in this section we choose $\textbf{M}$ to be diagonal, although we discuss other cases for the analysis 
of PAPER-64 data as explained in Section \ref{sec:Bias}. In a preview of what is to come, we also stress that the derivation that leads to Equation \eqref{eq:Wpplusbias} assumes that $\mathbf{R}$ and $\mathbf{x}$ are not correlated. If this assumption is violated, a simple application of the (now incorrect) formulae in this section can result in an improperly normalized power spectrum estimator that does not conserve power, i.e., one that has signal loss.
% JEA - is one really free to choose M and R independently?

@@

As a brief preview, we summarize our findings in the following sections here:
